Higher-Order unification for free!

### 2024-04-10 paper/main.tex
sections 3 and 4 are a draft

\noindent
The code above uses that possibility
in order to allocate space for the variables, i.e. sets their memory
address to \elpiIn{none} (a details not worth mentioning in the
previous sections).

@@

\subsubsection{Compilation of terms in \maybebeta}

@@

\noindent
In the base compilation scheme \elpiIn{progress1} is the identity
on both the links and the substitution, so the fixpoint trivially terminates.
Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
and justify why the don't hinder termination. For brevity we omit the code
that applies the substitution \elpiIn{S1} to all terms in \linkStore.

Since compilation moves problematic terms out of the sigh of \Ue{},
that procedure can only perform a partial occur check. For example the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if a $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: We don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$, hence
$f~Y$ and that fails the occur check.
The procedure \elpiIn{occur-check-links} is in charge of ensuring that
each link does not represent a (suspended) unification problem doomed
to fail because of occur check. This check is needed in order to
guarantee \fullref{prop:fidelity}.

@@

Decompiling the substitution requires to first force the
progress of links and then allocating new unassigned variables
in the substitution for \Fo{} and finally decompiling all
assignments. Note that \cref{inv:linklhs} and the
occur check allows us to update the subst.

\input{code/decompile}

\todo{commit links + complete-map + abs to lam}
% \noteE{What is commit-links and complete-mapping?}

@@

\begin{invariant}

  TODO: dire che il mapping è bijective
  \label{inv:map-bijective}
\end{invariant}

@@

The code given so far applies to terms in $\beta\eta$-normal form where
unification variables in \Fo{} can occur non linearly but always with
the same number of arguments, and where their arguments are distinct names
(as per \llambda).

\begin{lemma}[Compilation round trip] If
  \elpiIn{comp S T [] M [] _ [] _} then \elpiIn{decomp M T S}
\end{lemma}
\begin{proof}[Proof sketch]
trivial, since the terms are beta normal beta just builds an app. \noteD{Reformulate?}
\end{proof}

\begin{lemma}
\Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo above
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as
\Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can find this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the \Fo{} terms, the are two interesting cases:
\begin{itemize}
\item \elpiIn{fuva X ~\Uo~s}. In this case after \elpiIn{comp} we have
  $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
  $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
\item \elpiIn{fapp[fuva X|L] ~\Uo~s}. In this case
 we have $Y_{\vec{x}} \Ue t$ that succeeds with
 $\sigma = \{ \vec{y} \vdash Y \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
 is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
 Thanks to \ref{clause:beta1}
 $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
\end{itemize}
Since the mapping is a bijection occur check in \Ho{} corresponds to occur
check in \Fo{}.
\end{proof}

\begin{lemma} \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold
\end{lemma}
\begin{proof}[Proof sketch]
Since \elpiIn{progress1} is trivial \fstep and \hstep are the same, that is
in this
context where input terms are $\beta\eta$-normal and we disregard $\eta$-equivalence
\Ue is equivalent to \Uo.
\end{proof}

@@

\subsection{Detection}
\subsection{Compilation and decompilation}
\subsection{Progress}

% \subsection{Example}

@@

\begin{lemma}
  The introduction of \linketa guarantees \fullref{prop:w-preservation}
  \label{lemma:unif-wellb}
\end{lemma}

>>>>

\noindent
The code above uses that possibility
in order to allocate space for the variables, i.e. sets their memory
address to \elpiIn{none} (a details not worth mentioning in the
previous sections).
\todo{explain fold6}

@@

@@

\subsubsection{Progress}
In the base compilation scheme \elpiIn{progress1} is the identity
on both the links and the substitution, so the fixpoint trivially terminates.
Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
and justify why the don't hinder termination. For brevity we omit the code
that applies the substitution \elpiIn{S1} to all terms in \linkStore.

\subsubsection{Occur check}
Since compilation moves problematic terms out of the sigh of \Ue{},
that procedure can only perform a partial occur check. For example the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if a $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: We don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$, hence
$f~Y$ and that fails the occur check.
The procedure \elpiIn{occur-check-links} is in charge of ensuring that
each link does not represent a (suspended) unification problem doomed
to fail because of occur check. This check is needed in order to
guarantee \fullref{prop:fidelity}.

@@

Decompiling the substitution involves three steps.

First and foremost problematic terms stored in
\linkStore have to be moved back into the game.
Since links are of the form uvar = term (\fullref{inv:linklhs})
and are duplicate free (see dedup-beta dedup-eta),
one can turn a link X = t into an assignment $X \mapsto t$.
This can in general be achieved by unifying X with t. The case
where t is not in \llambda (link beta/llam) is discussed in section xx.

The second step amounts at allocating new variables in the
memory of \Fo{}. In particular some unif problems
such as $F x y = F x z$ requires to allocate a 
variable $G$ so that the assignment
$F_{ab} \mapsto G_a$ can be used to perform
required pruning.

The last step amounts at decompiling each assignment.
Decompiling a term is trivial. An assignment has an
abs node, as in move, can be eliminated by replacing
the bound variable by the actual term in scope.
In order to do this, one needs the \mapStore to be
a bijection. This is the job of \cref{sec:invariant1}.
\todo{polish}

\todo[inline]{dire che però si passa per una subst in cui ste abs
le cambio in lam. Nel codice Coq ci scrivevamo il tipo nella
arity, e quindi sappiamo fare i lambda bene, senza perdita di
informazione. Qui i lam non hanno info, facile. Ma in generale
bisogna spiegare come ci si salva. Ci dormo su: o non generiamo
la subst ma solo il primo termine (la query iniziale) istanziato
(funziona sempre, la prova è quella sopra) oppure bisogna siegare
tutto sto casino e serve un po' di spazio.}

%\input{code/decompile_type}

% \noteE{What is commit-links and complete-mapping?}

@@

@@

The code given so far still makes no use
of the higher order nature of the ML unif language,
indeed the scope of unif variables generated by the
compiler is always empty, so \Ue{} is first order.

Still, if \foUnifPb is already \wellb, we can set up a proof
that will also work when comp enforces \wellb and  hstep
preserves it, and when terms in \llambda are mapped to ho variables
with a scope.

\begin{lemma}[Compilation round trip] If
  \elpiIn{comp S T [] M [] _ [] _} then \elpiIn{decomp M T S}
\end{lemma}
\begin{proof}[Proof sketch]
trivial if the mapping is a bijection and 
the terms are beta normal.
some discussion about commit maybellam to be done later.
\end{proof}

\begin{lemma}
\Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo above
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as
\Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can find this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the \Fo{} terms is only one interesting cases:
\begin{itemize}
\item \elpiIn{fuva X ~\Uo~s}. In this case after \elpiIn{comp} we have
  $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
  $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
\end{itemize}
Since the mapping is a bijection occur check in \Ho{} corresponds to occur
check in \Fo{}.
\end{proof}

\begin{theorem}[Fidelity in \wellb] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold
\end{theorem}
\begin{proof}[Proof sketch]
Since \elpiIn{progress1} is trivial \fstep and \hstep are the same, that is
in this
context where input terms are $\beta\eta$-normal and all is \wellb,
\Ue is equivalent to \Uo.
\end{proof}

@@

\paragraph{Detection} trivial, pattern-fragment.
\subsection{Compilation and decompilation}

\paragraph{Decompilation} since no link, nothing new

\paragraph{Progress} since ML unif is complete here, no
need to move terms aside, just use uva is enough.

\begin{lemma}
  \Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo
  in \cref{sec:founif}
  \end{lemma}
  \begin{proof}[Proof sketch]
  If we look at the \Fo{} terms, the is one more case interesting cases:
  \begin{itemize}
  \item \elpiIn{fapp[fuva X|L] ~\Uo~s}. In this case
   we have $Y_{\vec{x}} \Ue t$ that succeeds with
   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
\todo{ste x escono da L}   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
   Thanks to \ref{clause:beta1}
   $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
  \end{itemize}
  \end{proof}

\begin{lemma}[\wellb-enforcement]\label{lem:w-enforcement-maybebeta} Even if $\foUnifPb \cap \maybebeta \ne \emptyset$,
  $\hoUnifPb \cup \maybebeta = \emptyset$
\end{lemma}
\begin{proof}[Proof sketch]
  problematic terms are mapped to uva by comp, the problematic fapp node is gone.
\end{proof}

\begin{theorem}[Fidelity in \maybebeta] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold
\end{theorem}
\begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
  it is the same as in \cref{sec:simulation}, even if now
  we really need \Ue{} to deal with \llambda, while before a FO unif
  would have done.
\end{proof}
% \subsection{Example}

@@

\begin{lemma}
  The introduction of \linketa guarantees \fullref{prop:w-preservation}
  \label{lemma:unif-wellb}
\end{lemma}
\todo{is this W-enforcing?}
